The effect of climate forcing on numerical simulations of the Cordilleran Ice Sheet at the Last Glacial Maximum

### 2013-07-01 sections/intro.tex
Some commenting and small changes in the intro

Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.

%Irina  I think it would be reasonable to discuss the limitations of GCM in this context

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
\julien[noline]{Should this literature list be extended to studies not covering North America?}

@@

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at the Last Glacial Maximum (LGM) reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a Parallel Ice Sheet Model \citep[PISM,][]{web:pism} \julien{Recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While \citet{robert-1991} modelling domain covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.

>>>>

Arguably the most physically sound way to force an ice sheet model in the past is to couple it with a General Circulation Model (GCM) \citep{yoshimori-etal-2001,calov-etal-2002,abeouchi-etal-2007,charbit-etal-2013}. However the computational demand of GCMs is such that only models of intermediate complexity can run on the time-scales of tens of thousands of years characteristic of ice sheet growth and decay.
\julien{Add words on limitations of GCM in this context?}

Climatologies obtained from uncoupled GCM palaeo-climate simulations such as produced within the PMIP project \citep{joussaume-taylor-1995} provide a more accurate representation of past climate and are commonly used to force ice sheet models. Because they are only available for specific periods of time, they require either an assumption of steady-state \citep{huybrechts-tsiobbel-1996}, or interpolation through time between climatologies from different periods, which can be linear \citep{charbit-etal-2002}, or modulated by a ``glacial index'' weighting function derived from ice core $\delta^{18}$\,O records \citep{marshall-clarke-1999,tarasov-peltier-2004,zweck-huybrechts-2005,gregoire-etal-2012}. An important drawback in this approach is that palaeo-climate simulations themselves rely on global ice sheet reconstructions such as the ICE-4G \citep{peltier-1994} for their surface topographic boundary condition. This indirectly controls subsequently modelled ice sheet geometries, whose spatial extent tend to conform to these reconstructions.
\julien[noline]{Should I include references not covering the Cordilleran Ice Sheet?}

@@

The Cordilleran Ice Sheet (Fig.~\ref{fig:locmap}) covered an area that presently experiences strong regional variations in climate. In a numerical modelling perspective, it is one of the least studied palaeo-ice sheets of the Northern Hemisphere, despite the fact that significant geomorphological data is available to constrain its extent \citep{jackson-clague-1991,dukrodkin-1999,kaufman-manley-2004,kleman-etal-2010,margold-etal-2011}. Although the Cordilleran Ice Sheet was not the primary target of these simulations, it has previously been modelled as part of efforts to model ice sheets in North America \citep{marshall-clarke-1999,calov-etal-2002,tarasov-peltier-1997,tarasov-peltier-2004,gregoire-etal-2012}, the Northern Hemisphere \citep{huybrechts-tsiobbel-1996,greve-etal-1999,charbit-etal-2002,charbit-etal-2007,charbit-etal-2013,johnson-fastook-2002,rodgers-etal-2004,bintanja-etal-2005,zweck-huybrechts-2005,abeouchi-etal-2007} or world-wide \citep{yoshimori-etal-2001}. While these studies reproduce the magnitude of North American glaciation at LGM reasonably well, there exist a tendency in simulations independent from ice sheet reconstructions such as the ICE-4G to predict excessive ice cover in parts of northern Yukon Territory and continental Alaska that have remained ice-free throughout the Pleistocene \citep{dukrodkin-1999,kaufman-manley-2004}.

Here we use a PISM, a Parallel Ice Sheet Model \citep{web:pism} \julien{Citation recommended by \url{http://www.pism-docs.org/wiki/doku.php?id=citing_pism}} to simulate the extent and thickness of Cordilleran Ice Sheet at the LGM. We force our model with multiple climate datasets and compare our results to the mapped LGM ice sheet margin from \citet{dyke-2004}. To our knowledge, this is the first modelling study which specifically focus on the Cordilleran Ice Sheet since the one by \citet{robert-1991}. While the modelling domain of \citet{robert-1991} covered only the southern part of the ice sheet, our model also includes higher resolution and an improved treatment of ice thermodynamics and bedrock response due to large advances made in ice sheet modelling since then. Model set-up is presented in section~\ref{sec:model} and climate forcing in section~\ref{sec:climate}. Results are exposed in section~\ref{sec:results} and discussed in section~\ref{sec:discussion}.
